A Short and Mechanized Consistency Proof for Dependent Type Theory

### 2024-02-14 paper/paper.tex
Add more text to related work

The glaring gap here is the
lack of fully dependently typed systems where computations may appear
at the type-level. In fact, it is far from obvious why proof by
logical relation is even applicable to dependent types, since the type
may very well-be a computation that is yet to be evaluated.
While it is assuring that proof by logical relations for dependent
types is available in mechanized forms, % the key to address the
% complexities of dependent types is obscured in
\citet{nbeincoq,decagda,martin-lof-a-la-coq} all involve relational,
Kripke-style models that obscure the technique for addressing
type-level computation. In fact, the added complexivity is evident
from the size of their code base. For example, \citet{nbeincoq}
involves 22 thousand lines of Coq code.

@@

\begin{figure}[h]
\begin{minipage}{0.9\textwidth}
\drules[P]{$[[a => b]]$}{Parallel Reduction}{Var, Set, Void, Pi, Abs, App, AppAbs, True, False, If, IfTrue, IfFalse,
  Bool, Eq, Refl, J, JRefl}
\drules[PS]{$[[a =>+ b]]$}{Transitive Closure of Parallel Reduction}{One, Step}
\drules[C]{$[[a <=> b]]$}{Coherence}{Intro}
\end{minipage}
\caption{Parallel reduction and coherence}
\label{fig:par}
\end{figure}

Figure~\ref{fig:par} shows the definition of the parallel reduction
relation, which takes the form $[[a => b]]$. We use $[[a =>+ b]]$ to
represent its transitive and reflexive closure, which in turn allows us to define
the coherence relation $[[a <=> b]]$.
We say that two terms $[[a]]$
and $[[b]]$ are coherent if they can eventually reduce to some common
term $[[c]]$ through parallel reduction. The symmetric notation of
coherence suggests that it is an equivalence relation.% , and in fact, we
% will use coherence as the equational theory for our type conversion
% rule.

@@

Unfortunately, both versions not only violate the syntactic strict
positivity constraint required in proof assistants, but
are in fact genuinely not monotone over the domain of
relations.
Intuitively, the failure of monotonicity stems from the fact
that the witness picked in the precondition is not necessarily the
same witness being referred to in the post condition. While it might
be possible to restrict the domain with additional constraints such as
functionality and inversion properties, we opt for our current
skolemized formulation of \rref{I-Pi} so we immediately obtain a
well-defined inductive relation and a usable induction principle. The
more intuitive introduction rules can be later shown to be admissible
a posteriori.

@@

\section{Related Work}
\label{sec:relatedwork}
\citet{Martin-Lof-1973}, \citet{geuvers1994short}, and
\citet{barendregt:lambda-calculi-with-types} are some of the earlier
works that establish metatheoretic results strong enough to derive
consistency for dependently typed systems. Some of these techniques
are still in use in more recent works. For example, the technique for
proving strong normalization by \citet{geuvers1994short} is adopted by
\citet{moon2021graded} to show the same property to a dependently
typed system extended with modalities.

More recently, there have been efforts at mechanizing consistency
or decidability of type conversion proofs using proof assistants such
as Coq and Agda.

>>>>

The glaring gap here is the
lack of fully dependently typed systems where computations may appear
at the type-level. It is far from obvious why proof by
logical relation is even applicable to dependent types, since the type
may very well-be a computation that is yet to be evaluated.
While it is assuring that proof by logical relations for dependent
types is available in mechanized forms, % the key to address the
% complexities of dependent types is obscured in
\citet{nbeincoq,decagda,martin-lof-a-la-coq} all involve relational,
Kripke-style models that obscure the technique for addressing
type-level computation. The added complexivity is evident
from the size of their code base. For example, \citet{nbeincoq}
involves 22 thousand lines of Coq code.

@@

\begin{figure}[h]
\begin{minipage}{0.9\textwidth}
\drules[P]{$[[a => b]]$}{Parallel Reduction}{Var, Set, Void, Pi, Abs, App, AppAbs, True, False, If, IfTrue, IfFalse,
  Bool, Eq, Refl, J, JRefl}
\drules[PS]{$[[a =>+ b]]$}{Transitive Closure of Parallel Reduction}{Refl, Step}
\drules[C]{$[[a <=> b]]$}{Coherence}{Intro}
\end{minipage}
\caption{Parallel reduction and coherence}
\label{fig:par}
\end{figure}

Figure~\ref{fig:par} shows the definition of the parallel reduction
relation, which takes the form $[[a => b]]$. We use $[[a =>+ b]]$ to
represent its transitive and reflexive closure, which in turn allows us to define
the coherence relation $[[a <=> b]]$.
We say that two terms $[[a]]$
and $[[b]]$ are coherent if they can eventually reduce to some common
term $[[c]]$ through parallel reduction. The symmetric notation of
coherence suggests that it is an equivalence relation.

@@

Unfortunately, both versions not only violate the syntactic strict
positivity constraint required in proof assistants, but
are genuinely non-monotone over the domain of
relations.
Intuitively, the failure of monotonicity stems from the fact
that the witness picked in the precondition is not necessarily the
same witness being referred to in the post condition. While it might
be possible to restrict the domain with additional constraints such as
functionality and inversion properties, we opt for our current
skolemized formulation of \rref{I-Pi} so we immediately obtain a
well-defined inductive relation and a usable induction principle. The
more intuitive introduction rules can be later shown to be admissible
a posteriori.

@@

\section{Related Work}
\label{sec:relatedwork}
\citet{Martin-Lof-1973}, \citet{geuvers1994short}, and
\citet{barendregt:lambda-calculi-with-types} are some of the earlier
works that establish metatheoretic results strong enough to derive
consistency for dependently typed systems. Some of these techniques
are still in use in more recent works. For example, the technique for
proving strong normalization by \citet{geuvers1994short} is adapted by
\citet{moon2021graded} to show the same property to a dependently
typed system extended with modalities.

\citet{decagda} mechanizes in Agda the decidability of type
conversion rule for a dependently typed language with one predicative
universe level and typed judgmental equality with function
$\eta$-law. Unlike our logical relation for \lang{}, \citet{decagda}
uses a Kripke-style logical relation parameterized over an
type-directed equivalence relation satisfying certain properties. We
are able to avoid the complexivity of a Kripke-style logical relation
by defining our logical relation in terms of untyped parallel
reduction, which does not impose any scoping constraint on the
terms. \citet{decagda} defines their logical relation using the
induction-recursion scheme, which is available in Agda but not in Coq.
\citet{martin-lof-a-la-coq} manages to encode the logical relation
from \citet{decagda} in the predicative fragment of Coq using a
special way of encoding induction recursion. Their work further
extends the decidability of type conversion result from~\citet{decagda} to the decidability
type checking of a bidrectional type system.

\citet{anand2014towards} mechanizes the metatheory of
Nuprl~\citep{constable1986implementing} in Coq. The metatheory is an
extensional type theory with features such as dependent functions,
inductive types, and a full universe hierarchy. \citet{nbeincoq}
mechanizes the normalization-by-evaluation algorithm in Coq for a
dependently typed language with one predicative universe, similar to
\citet{decagda} and \citet{martin-lof-a-la-coq}. Both
\citet{anand2014towards} and \citet{nbeincoq} leverage the
impredicative \texttt{Prop} sort of Coq to define the interpretation
of dependent function types and thus are closely related to our
mechanization. However, instead of explaining the purely inductive
logical relation as a convoluted workaround to the lack of
induction-recursion in Coq, we give a self-contained explanation of
our logical relation.

Finally, it is worth pointing out that all systems we have discussed
so far builds a relational model for their logical relation. Recall
that our logical relation takes the form $[[InterpR i A S]]$ where
$[[S]]$ is a set of terms. In a relational model, each type is
interpreted as a partial equivalence relation over terms rather than a
set. A relational model is necessary for an extensional type theory
such as Nuprl, though \citet{nbeincoq,decagda,martin-lof-a-la-coq} all
use a relational model either to justify the function $\eta$-law or to
derive $\Pi$-injectivity. In Section~\ref{sec:extension}, our $\eta$
law for function is baked into the untyped reduction
relation and therefore a relational model is not needed.

The dependently typed systems we have discussed so far also vary
greatly in expressiveness. For example, the Calculus of
Constructions~\citep{CoC} and MLTT without universe levels are unable
to encode large elimination. \citet{decagda} and \citet{nbeincoq} lack
identity types. \citet{nbeincoq} lacks $\eta$-law for functions. Among
the mechanized systems we have discussed so far, only
\citet{anand2014towards} has a full universe hierarchy.
Our object language \lang{}, while small, has a decent coverage of
features commonly seen in dependently typed languages, including
dependent pattern matching, dependent function types, a full universe
hierarchy, an intensional identity type. We leave features such as
type-directed reduction (necessary for unit $\eta$-law) and
impredicative sorts as part of our future work.
